class: part-slide

# Introduction à <i class="fab fa-r-project"></i>

---

# Installation de <i class="fab fa-r-project"></i>

Site internet officiel&nbsp;: https://www.r-project.org/

> R is a language and environment for statistical computing and graphics.
> It is a GNU project which is similar to the S language and environment which was developed at Bell Laboratories (formerly AT&T, now Lucent Technologies) by John Chambers and colleagues.
> R can be considered as a different implementation of S.
> There are some important differences, but much code written for S runs unaltered under R.
>
> R provides a wide variety of statistical (linear and nonlinear modelling, classical statistical tests, time-series analysis, classification, clustering, ...) and graphical techniques, and is highly extensible.
> The S language is often the vehicle of choice for research in statistical methodology, and R provides an Open Source route to participation in that activity.
>
> One of R’s strengths is the ease with which well-designed publication-quality plots can be produced, including mathematical symbols and formulae where needed.
> Great care has been taken over the defaults for the minor design choices in graphics, but the user retains full control.

.center.font200[&rArr; [**Download R**](https://cloud.r-project.org/)]

---

# Sauvegarder le code ... Pas l'environnement&nbsp;!

.pull-left.font150[

Les scripts&nbsp;:

* "*standards*" dans un fichier `.R`

* "*Rmarkdown*" dans un fichier `.Rmd`

]

.pull-right.font150[

Les "données"&nbsp;:

* un objet dans un `.rds`

* une liste d'objet dans un `.Rdata`

]

---

# Utiliser un IDE ("Integrated Development Environment")

.font150[

* RStudio&nbsp;:
    https://www.rstudio.com/products/rstudio/download/#download

* Emacs + ESS&nbsp;:
    https://ess.r-project.org

* vim + Nvim-R&nbsp;:
    https://medium.com/free-code-camp/turning-vim-into-an-r-ide-cd9602e8c217

* Visual Studio Code + vscode-R&nbsp;:
    https://github.com/REditorSupport/vscode-R

]

---

# Faire *table rase* à chaque démarrage&nbsp;!

.font150[

Lorsque vous quittez <i class="fab fa-r-project"></i>&nbsp;:

* Ne pas enregistrer votre espace de travail (*workspace*&ast;)&nbsp;!

Lorsque vous démarrez <i class="fab fa-r-project"></i>&nbsp;:

* Ne pas charger l'espace de travail sauvegardé précédemment&nbsp;!

]

.footnote.font150[
&ast; Le *workspace* est automatiquement sauvegardé/chargé via un fichier nommé `.Rdata` (là où s'exécute <i class="fab fa-r-project"></i>).
]

---

# Faire *table rase*&nbsp;! Dans Windows

.center[<img src = "content/media/rwindows.png" />]

---

# Faire *table rase*&nbsp;! Dans RStudio (*Tools > Global options*)

.center[<img src = "content/media/rstudio.png" />]

# Faire *table rase*&nbsp;!

.pull-left.font150[
+ Dans un Terminal

```{bash, eval = FALSE}
R --no-save --no-restore-data
```

]

.pull-right.font150[
+ `.bash_profile` / `.bashrc` / ...

```{bash, eval = FALSE}
alias R='R --no-save --no-restore-data'
```

]

---

#  Comment Réinitialiser <i class="fab fa-r-project"></i>&nbsp;? Avec `rm(list = ls())`&nbsp;?

.font150[

> If the first line of your R script is
>
> `rm(list = ls())`
>
> I will come into your office and SET YOUR COMPUTER ON FIRE 🔥.
>
> --- Jenny Bryan

]

---

# Que ne fait pas `rm(list = ls())`&nbsp;?

.font150[

* Réinitialiser la session <i class="fab fa-r-project"></i> en cours

* Réinitialiser les `options()`
    &#x2192; `options(stringsAsFactors = FALSE)`&ast;

* Réinitialiser le répertoire de travail
    `getwd()` &#x2192; `setwd()`

* Réinitialiser les extensions <i class="fab fa-r-project"></i> attachées

]

.footnote.font150[
&ast; Valeur par défaut <i class="fab fa-r-project"></i> v4.0.0 ([Kurt Hornik: stringsAsFactors](https://developer.r-project.org/Blog/public/2020/02/16/stringsasfactors/)).
]

---

# Mais alors que fait `rm(list = ls())`&nbsp;?

```{r, eval = FALSE, echo = TRUE}
help(rm)
```

.center[<img src = "content/media/rm-doc.png" />]

---

# Réinitialiser <i class="fab fa-r-project"></i>

.font150[

* Dans une console <i class="fab fa-r-project"></i>

    * Via le raccourci `Ctrl+D`

    * Via la fonction `q()` dans une console <i class="fab fa-r-project"></i>

]

---

# Réinitialiser <i class="fab fa-r-project"></i>

.pull-left.font150[

* Dans RStudio&nbsp;:

    * Via le raccourci `Ctrl+Shift+F10`

    * Via le menu
]

.pull-right[
  .center[<img src = "content/media/restartr.png" />]
]

---

# Désactiver les fichiers de démarrage

.font150[

* `--vanilla` - pour désactiver le chargement de **tous les fichiers** de démarrage

* `--no-init-file` - pour désactiver le chargement du fichier `.Rprofile`

* `--no-environ` - pour désactiver le chargement du fichier `.Renviron`

]

---

#  `setwd(...)` en première ligne d'un script ...

.font150[

> If the first line of your R script is
>
> `setwd("C:\Users\jenny\path\that\only\I\have")`
>
> I will come into your office and SET YOUR COMPUTER ON FIRE 🔥.
>
> --- Jenny Bryan

]

---

# Faisons le Point sur `setwd()`&nbsp;!

.pull-left.code150[

```{r, echo = TRUE, eval = FALSE}
library(ggplot2)

setwd("/path/to/a/directory/on/my/laptop/data")

df <- read.delim("data.csv")

p <- ggplot(df, aes(x, y)) + geom_point()

ggsave("../figs/scatterplot.png")
```

]

.pull-right.code150[

```{r, echo = TRUE, eval = FALSE}
library(ggplot2)

setwd("/path/to/a/directory/on/my/laptop/data")

df <- read.delim("data.csv")

p <- ggplot(df, aes(x, y)) + geom_point()

setwd("/path/to/a/directory/on/my/laptop/figs")

ggsave("scatterplot.png")
```

]

---

# L'extension <i class = "fab fa-r-project"></i> `here`

```{r, echo = FALSE}
if (!file.exists(here("content", "media", "here_ahorst.png"))) {
  download.file(
    url = "https://github.com/allisonhorst/stats-illustrations/raw/master/rstats-artwork/here.png",
    destfile = here("content", "media", "here_ahorst.png")
  )
}
```

.pull-left[
.center[<img src = "content/media/here_ahorst.png" />]
]

.pull-right.code60[

```{r}
here::here()
here::i_am("radvanced.Rmd")
here::i_am("content/part-0.Rmd")
fs::dir_tree(here::here(), recurse = FALSE)
```

]

---

# Utiliser un Mode "Projet" via un IDE

.font150[

* Démarrer un projet A&nbsp;:

    1. Démarre R.

    2. Défini le répetoire de travail de R comme la racine du projet A.

* Passer d'un projet A à un project B&nbsp;:

    1. Redémarre R.

    2. Défini le répetoire de travail de R comme la racine du projet B.

]

---

# Utiliser un Mode "Projet"

.font150[

* [__"What They Forgot to Teach You About R"__](https://rstats.wtf/project-oriented-workflow.html) --- Jenny Bryan & Jim Hester

* [__"Project-oriented Workflow"__](https://www.tidyverse.org/articles/2017/12/workflow-vs-script/) --- Jenny Bryan

]

---

# Lectures Utiles

.font150[

* [__"Code Smells and Feels"__](https://www.youtube.com/watch?v=7oyiPBjLAWY) --- Jenny Bryan

* [__"Advanced R"__](https://adv-r.hadley.nz/) --- Hadley Wickham

* [__"Project-oriented Workflow"__](https://www.tidyverse.org/articles/2017/12/workflow-vs-script/) --- Jenny Bryan

* [__"R for Data Science"__](https://r4ds.had.co.nz/) --- Garrett Grolemund & Hadley Wickham

* [__"What They Forgot to Teach You About R"__](https://rstats.wtf/) --- Jenny Bryan & Jim Hester

]
